---
title: "MAT1856/APM466 Assignment 1"
author: Yonglin Liu, 1004877207
date: February, 20
output: pdf_document
---

## Fundamental Questions - 25 points

### 1.

##### (a).

Governments simply printing money could cause inflation, while governments could easily borrow money by issuing bonds.

##### (b).

It might be because the market expects inflation to fall, or the market predicts the future economy will slow down so that the demand of long-term bonds increases.

##### (c).

The Fed increased asset purchases like Treasuries, then the demand for treasuries went up, then the yields went down, thus the risk free rate decreased to near 0, therefore, businesses and consumers tend to increase their spending which stimulates the economy.

### 2.

The 11 bonds I selected are: “CAN 0.5 Mar 1”, “CAN 0.25 Aug 1(2020)”, “CAN 1.75 Mar 1”, “CAN 0.25 Aug 1(2021)”, “CAN 2.25 Mar 1”, “CAN 1.5 Sep 1”, “CAN 1.25 Mar 1(2025)”, “CAN 0.5 Sep 1”, “CAN 0.25 Mar 1”, “CAN 1 Sep 1”, “CAN 1.25 Mar 1(2027)”.
To perform bootstrapping and interpolating, the maturity date of these bonds are about {less than 0.5 years, less than 1 years but greater than 0.5 years, less than 1.5 years but greater than 2 years, … , less than 5.5 years but greater than 5 years} from now respectively. All of these bonds have a term (time length from issue date to maturity date) of about 5 years, except two bonds have a term of around 2 years.

### 3.

The eigenvectors, which also called principle components, are particular linear combination of the stochastic process. Geometrically, the eigenvectors represent the direction of a new coordinate whose variance are maximized. The eigenvalues explained the proportion of total population variance due to the corresponding eigenvectors, which determine the magnitudes of the eigenvectors. For instance, the first eigenvector which has the largest eigenvalue is the linear combination of stochastic process with maximum variance.

## Empirical Questions - 75 points

### 4.

##### (a).

```{r, echo=F, message=F, warning=F}
# library
if (!require("jrvFinance")) install.packages("jrvFinance")
library(jrvFinance)
if (!require("lubridate")) install.packages("lubridate")
library(lubridate)
if (!require("tinytex")) install.packages("tinytex")
library(tinytex)

# hardcode all the data
ISIN <- c("CA135087G328", "CA135087L286", "CA135087H490",
          "CA135087M359", "CA135087J546", "CA135087J967",
          "CA135087K528", "CA135087K940", "CA135087L518",
          "CA135087L930", "CA135087M847")
Coupon <- c(0.50, 0.25, 1.75, 0.25, 2.25, 1.50, 1.25, 0.50, 
            0.25, 1.00, 1.25)
Issue_date <- as.Date(c("2016-10-10", "2020-05-03", "2017-10-05",
                        "2021-05-13", "2018-10-04", "2019-04-04",
                        "2019-10-10", "2020-04-02", "2020-10-08",
                        "2021-04-15", "2021-10-14")) 
Maturity_date <- as.Date(c("2022-03-01", "2022-08-01", "2023-03-01",
                           "2023-08-01", "2024-03-01", "2024-09-01",
                           "2025-03-01", "2025-09-01", "2026-03-01",
                           "2026-09-01", "2027-03-01"))
Close_Price_date <- as.Date(c("2022-01-10", "2022-01-11",
                              "2022-01-12","2022-01-13",
                              "2022-01-14","2022-01-17",
                              "2022-01-18", "2022-01-19",
                              "2022-01-20", "2022-01-21"))
day1 <- c(100.04, 99.8, 100.94, 98.808, 102.31, 100.56, 99.64,
          96.66, 95.11, 97.66, 98.41)
day2 <- c(100.03, 99.79, 100.93, 98.8, 102.31, 100.63, 99.67,
          96.72, 95.17, 97.7, 98.475)
day3 <- c(100.03, 99.78, 100.91, 98.781, 102.26, 100.62, 99.61,
          96.77, 95.12, 97.67, 98.4)
day4 <- c(100.03, 99.77, 100.88, 98.763, 102.22, 100.58, 99.59,
          96.77, 95.14, 97.695, 98.44)
day5 <- c(100.02, 99.76, 100.86, 98.736, 102.18, 100.52, 99.49,
          96.68, 95.04, 97.585, 98.31)
day6 <- c(100.02, 99.72, 100.76, 98.61, 101.99, 100.29, 99.25,
           96.43, 94.75, 97.24, 97.955)
day7 <- c(100, 99.7, 100.71, 98.553, 101.91, 100.21, 99.12, 96.26,
          94.53, 97.02, 97.7)
day8 <- c(100, 99.7, 100.71, 98.549, 101.9, 100.15, 99.11, 96.15,
          94.5, 96.96, 97.66)
day9 <- c(100, 99.72, 100.71, 98.558, 101.88, 100.17, 99.1, 96.19,
          94.52, 97, 97.685)
day10 <- c(100.01, 99.74, 100.75, 98.611, 101.97, 100.27, 99.26,
           96.31, 94.7, 97.225, 97.94)
# the dataset for the 11 bonds include all the information we need
Bond <- cbind(ISIN, Coupon, Issue_date, Maturity_date, day1, day2,
              day3, day4, day5, day6, day7, day8, day9, day10)
# the matrix include all the close price in 10 days for 11 bonds
Close_Price <- matrix(c(day1, day2, day3, day4, day5, day6, day7, 
                      day8, day9, day10), nrow = 10, ncol = 11, 
                      byrow = T)
```

By using formula $P(B)=\sum_ip_ie^{-r(B)*t_i}$, we can get the YTM for all 11 bonds in 10 days, where P(B) is the dirty price and r(B) is the YTM. Then, plotting the yield curve we get the Figure 1, where each line represent the yield curve of each day. \

In order to compute YTM curve at time to maturity {0.5,1,1.5, 2,...,5} years, I will use linear interpolation $r=r_1+\frac{(r_2-r_1)}{t_2-t_1}*(t-t_1)$ where $t_1<t<t_2$. Then we can get the yield curve with linear interpolation in Figure 2.

```{r, echo=F, message=F, warning=F, out.width="80%", out.height="80%"}
# compute YTM matrix
ytm <- matrix(rep(0, 110), nrow = 10, ncol = 11, byrow = T)
for (j in 1:11) {
  for (i in 1:10) {
    ytm[i,j] <- bond.yield(settle = Close_Price_date[i], 
                           mature = Maturity_date[j], 
                           coupon = Coupon[j]/100, freq = 2, 
                           price = Close_Price[i,j], 
                           convention = "ACT/ACT", 
                           redemption_value = 100)
  }
}

# compute time to Maturity 
time_yr <- matrix(rep(0, 110), nrow = 10, ncol = 11, byrow = T)
for (j in 1:11) {
  for (i in 1:10) {
    time_yr[i,j] <- time_length(
      interval(ymd(Close_Price_date[i]), ymd(Maturity_date[j])),
      "year")
  }
}

# plot the YTM curve
par(mfrow=c(1,2))
color_line = c("palegreen","yellowgreen", "seagreen", "royalblue",
               "skyblue","slateblue", "purple", "purple4", "black")
plot(time_yr[1,], ytm[1,], type = "o", main = "Figure 1: YTM Curve",
     xlab = "time to maturity (in years)", ylab = "YTM",
     col="yellow3", ylim = c(0.002, 0.0175))
for (i in 2:10) {
  lines(time_yr[i,], ytm[i,], col=color_line[i-1], type = "o")
}
legend("bottomright", pch = c(15,15), 
       legend = c("day1", "day2","day3", "day4", "day5", "day6",
                  "day7", "day8", "day9", "day10"),
       col=c("yellow3", color_line), lty=1.3, cex=0.6)

# estimating YTM using linear interpolation
years <- c(0.5, 1, 1.5, 2, 2.5, 3, 3.5, 4, 4.5, 5)
ytm_int <- matrix(rep(0,100), nrow=10, ncol=10, byrow = T)
for (i in 1:10) {
  for (j in 1:10) {
    est_rate = ytm[i,j]+((ytm[i,(j+1)]-ytm[i,j])/
      (time_yr[i,(j+1)]-time_yr[i,j]))*(years[j]-time_yr[i,j])
    ytm_int[i,j]=est_rate
  }
}

# plot estimated YTM using linear interpolation
plot(years, ytm_int[1,], type = "o", 
     main = "Figure 2: YTM Curve 
     with Linear Interpolation", 
     xlab = "time to maturity (in years)", 
     ylab = "YTM (estimated)", col="yellow3",
     ylim = c(0.002, 0.0175))
for (i in 2:10) {
  lines(years, ytm_int[i,], col=color_line[i-1], type = "o")
}
legend("bottomright", pch = c(15,15), 
       legend = c("day1", "day2","day3", "day4", "day5", "day6",
                  "day7", "day8", "day9", "day10"),
       col=c("yellow3", color_line), lty=1.3, cex=0.6)

```

##### (b).

By using formula $P=\sum_i^np_ie^{-r(t_i)*t_i}$ and bootstrapping method $P=\sum_i^{n-1}p_ie^{-r(t_i)*t_i}+p_ne^{-r(t_n)*t_n}$, where $t_i$ for $i<n$ is known and $t_n$ is unknown, we compute the spot rate at each $t_n$. Recursively, we can get the spot curve in Figure 3.

```{r, echo=F, message=F, warning=F, out.width="80%", out.height="80%"}
# computing dirty price(=clean price+accrued interest)
dirty_price <- matrix(rep(0, 110), nrow = 10, ncol = 11, byrow = T)
for (j in 1:11) {
  for (i in 1:10) {
    dirty_price[i,j]=Close_Price[i,j]+
      bond.TCF(settle = Close_Price_date[i],
               mature = Maturity_date[j], 
               coupon = Coupon[j]/100, freq = 2, 
               convention = "ACT/ACT", 
               redemption_value = 100)$accrued
  }
}

#computing cashflow
cf1=c(100.25)
cf2=c(0.125, 100.125)
cf3=c(0.875, 0.875, 100.875)
cf4=c(rep(0.125, 3), 100.75)
cf5=c(1.125, 1.125, 1.125, 1.125, 101.125)
cf6=c(0.75, 0.75, 0.75, 0.75, 0.75, 100.75)
cf7=c(0.625, 0.625, 0.625, 0.625, 0.625, 0.625, 100.625)
cf8=c(rep(0.25, 7), 100.25)
cf9=c(rep(0.125, 8), 100.125)
cf10=c(rep(0.5, 9), 100.5)
cf11=c(rep(0.625, 10), 100.625)

#computing spot rate
spot <- matrix(rep(0, 110), nrow = 10, ncol = 11, byrow = T)
for (i in 1:10) {
  t1=time_yr[i,1]
  f1=function(x) dirty_price[i,1]-cf1[1]*exp(-x*t1)
  s1=uniroot(f1, c(0,1))$root
  
  t2=time_yr[i,2]
  f2=function(x) dirty_price[i,2]-cf2[1]*exp(-s1*t1)-
    cf2[2]*exp(-x*t2)
  s2=uniroot(f2, c(0,1))$root
  
  t3=time_yr[i,3]
  f3=function(x) dirty_price[i,3]-cf3[1]*exp(-s1*t1)-
    cf3[2]*exp(-s2*t2)-cf3[3]*exp(-x*t3)
  s3=uniroot(f3, c(0,1))$root
  
  t4=time_yr[i,4]
  f4=function(x) dirty_price[i,4]-cf4[1]*exp(-s1*t1)-
    cf4[2]*exp(-s2*t2)-cf4[3]*exp(-s3*t3)-cf4[4]*exp(-x*t4)
  s4=uniroot(f4, c(0,1))$root
  
  t5=time_yr[i,5]
  f5=function(x) dirty_price[i,5]-cf5[1]*exp(-s1*t1)-
    cf5[2]*exp(-s2*t2)-cf5[3]*exp(-s3*t3)-cf5[4]*exp(-s4*t4)-
    cf5[5]*exp(-x*t5)
  s5=uniroot(f5, c(0,1))$root
  
  t6=time_yr[i,6]
  f6=function(x) dirty_price[i,6]-cf6[1]*exp(-s1*t1)-
    cf6[2]*exp(-s2*t2)-cf6[3]*exp(-s3*t3)-cf6[4]*exp(-s4*t4)-
    cf6[5]*exp(-s5*t5)-cf6[6]*exp(-x*t6)
  s6=uniroot(f6, c(0,1))$root
  
  t7=time_yr[i,7]
  f7=function(x) dirty_price[i,7]-cf7[1]*exp(-s1*t1)-
    cf7[2]*exp(-s2*t2)-cf7[3]*exp(-s3*t3)-cf7[4]*exp(-s4*t4)-
    cf7[5]*exp(-s5*t5)-cf7[6]*exp(-s6*t6)-cf7[7]*exp(-x*t7)
  s7=uniroot(f7, c(0,1))$root
  
  t8=time_yr[i,8]
  f8=function(x) dirty_price[i,8]-cf8[1]*exp(-s1*t1)-
    cf8[2]*exp(-s2*t2)-cf8[3]*exp(-s3*t3)-cf8[4]*exp(-s4*t4)-
    cf8[5]*exp(-s5*t5)-cf8[6]*exp(-s6*t6)-cf8[7]*exp(-s7*t7)-
    cf8[8]*exp(-x*t8)
  s8=uniroot(f8, c(0,1))$root
  
  t9=time_yr[i,9]
  f9=function(x) dirty_price[i,9]-cf9[1]*exp(-s1*t1)-
    cf9[2]*exp(-s2*t2)-cf9[3]*exp(-s3*t3)-cf9[4]*exp(-s4*t4)-
    cf9[5]*exp(-s5*t5)-cf9[6]*exp(-s6*t6)-cf9[7]*exp(-s7*t7)-
    cf9[8]*exp(-s8*t8)-cf9[9]*exp(-x*t9)
  s9=uniroot(f9, c(0,1))$root
  
  t10=time_yr[i,10]
  f10=function(x) dirty_price[i,10]-cf10[1]*exp(-s1*t1)-
    cf10[2]*exp(-s2*t2)-cf10[3]*exp(-s3*t3)-cf10[4]*exp(-s4*t4)-
    cf10[5]*exp(-s5*t5)-cf10[6]*exp(-s6*t6)-cf10[7]*exp(-s7*t7)-
    cf10[8]*exp(-s8*t8)-cf10[9]*exp(-s9*t9)-cf10[10]*exp(-x*t10)
  s10=uniroot(f10, c(0,1))$root
  
  t11=time_yr[i,11]
  f11=function(x) dirty_price[i,11]-cf11[1]*exp(-s1*t1)-
    cf11[2]*exp(-s2*t2)-cf11[3]*exp(-s3*t3)-cf11[4]*exp(-s4*t4)-
    cf11[5]*exp(-s5*t5)-cf11[6]*exp(-s6*t6)-cf11[7]*exp(-s7*t7)-
    cf11[8]*exp(-s8*t8)-cf11[9]*exp(-s9*t9)-cf11[10]*exp(-s10*t10)-
    cf11[11]*exp(-x*t11)
  s11=uniroot(f11, c(0,1))$root
  spot[i,1]=s1
  spot[i,2]=s2
  spot[i,3]=s3
  spot[i,4]=s4
  spot[i,5]=s5
  spot[i,6]=s6
  spot[i,7]=s7
  spot[i,8]=s8
  spot[i,9]=s9
  spot[i,10]=s10
  spot[i,11]=s11
}

# plot spot curve
par(mfrow=c(1,2))
plot(time_yr[1,], spot[1,], type = "o", 
     main = "Figure 3: Spot Curve", 
     xlab = "time to maturity (in years)", ylab = "spot rate",
     col="yellow3", ylim = c(0.002, 0.0175))
for (i in 2:10) {
  lines(time_yr[i,], spot[i,], col=color_line[i-1], type = "o")
}
legend("bottomright", pch = c(15,15), 
       legend = c("day1", "day2","day3", "day4", "day5", "day6",
                  "day7", "day8", "day9", "day10"),
       col=c("yellow3", color_line), lty=1.3, cex=0.6)

# part (c)
# estimating spot rate using linear interpolation
years <- c(0.5, 1, 1.5, 2, 2.5, 3, 3.5, 4, 4.5, 5)
spot_int <- matrix(rep(0,100), nrow=10, ncol=10, byrow = T)
for (i in 1:10) {
  for (j in 1:10) {
    est_spot = spot[i,j]+((spot[i,(j+1)]-spot[i,j])/
      (time_yr[i,(j+1)]-time_yr[i,j]))*(years[j]-time_yr[i,j])
    spot_int[i,j]=est_spot
  }
}

# get the estimated spot rate we need to compute forward rate
spot_for <- matrix(rep(0, 50), nrow=10, ncol = 5)
for (j in 1:5) {
  spot_for[,j]=spot_int[,2*j]
}

# compute 1-year forward curve
forward_rate <- matrix(rep(0, 40), nrow=10, ncol = 4)
for (i in 1:10) {
  for (j in 1:4) {
    forward_rate[i,j] = (spot_for[i,j+1]*(j+1)-spot_for[i,1]) /j
  }
}

# plotting the forward curve
plot(c(2,3,4,5), forward_rate[1,], type = "o", 
     main = "Figure 4: Forward Curve", 
     xlab = "time to maturity (in years)", 
     ylab = "Forward Rate", col="yellow3",
     ylim = c(0.015, 0.02))
for (i in 2:10) {
  lines(c(2,3,4,5), forward_rate[i,], col=color_line[i-1], 
        type = "o")
}
legend("topleft", pch = c(15,15), 
       legend = c("day1", "day2","day3", "day4", "day5", "day6",
                  "day7", "day8", "day9", "day10"),
       col=c("yellow3", color_line), lty=1, cex=0.5)

```

##### (c).

Similarly to part 4(a), we can use linear interpolation and get the spot curve with spot rate at time {1,2,3,4,5} years. Then applying the value to $r_{12}=-\frac{logP(t,T_2-logP(t,T_2))}{T_2-T_1}=\frac{r_2(T_2-t)-r_1(T_1-t)}{T_2-T_1}$, we can get the forward curve in Figure 4.


### 5.

The covariance matrix for daily log-returns of yield is: 
$$
\begin{pmatrix}
0.001221 & 0.000945 & 0.000832 & 0.000743 & 0.000723 \\
0.000945 & 0.000766 & 0.000688 & 0.000609 & 0.000595 \\
0.000832 & 0.000688 & 0.000652 & 0.000589 & 0.000570  \\
0.000743 & 0.000609 & 0.000589 & 0.000567 & 0.000540 \\
0.000723 & 0.000595 & 0.000570 & 0.000540 & 0.000524
\end{pmatrix}
$$

```{r, echo=F, message=F, warning=F, include=F}
# computing log_YTM with YTM with linear interpolation
log_YTM <- matrix(nrow=9, ncol=5)
for (i in 1:9) {
  for (j in 1:5) {
    log_YTM[i,j] <- log(ytm_int[(i+1), (2*j)]/ytm_int[i,(2*j)])
  }
}


# covariance matrix for log(YTM)
cov_ytm <- cov(log_YTM)
cov_ytm
```

The covariance matrix for daily log-returns of forward rate is:
$$
\begin{pmatrix}
 0.0004110427 & 0.0004583211 & 0.0004024626 & 0.0004015246\\
 0.0004583211 & 0.0005566225 & 0.0005063865 & 0.0004919808\\
 0.0004024626 & 0.0005063865 & 0.0005048462 & 0.0004791171 \\
 0.0004015246 & 0.0004919808 & 0.0004791171 & 0.0004671179
\end{pmatrix}
$$


```{r, echo=F, message=F, warning=F, include=F}
# computing log_forward
log_forward <- matrix(nrow=9, ncol=4)
for (i in 1:9) {
  for (j in 1:4) {
    log_forward[i,j] <- 
      log(forward_rate[i+1, j]/forward_rate[i,j])
  }
}

# covariance matrix for log(YTM)
cov_fwd <- cov(log_forward)
cov_fwd
```

### 6.

The eigenvalues for covariance matrix of log yield are $e_{y1}=0.0035680$, $e_{y2}=0.0001281$, $e_{y3}=0.0000242$, $e_{y4}=0.0000056$, and $e_{y5}=0.0000034$, and the eigenvectors are $$v_{y1}=(-0.5706, -0.4587, -0.4215, -0.3843, -0.3723)^T$$ $$v_{y2}=(0.6518, 0.2155, -0.2530, -0.5172, -0.4441)^T$$ 
$$v_{y3}=(-0.4377, 0.5568, 0.5388, -0.4082, -0.2037)^T$$ 
$$v_{y4}=(0.1938, -0.5336, 0.6303, 0.1498, -0.5079)^T$$
$$v_{y5}=(0.1429, -0.3853, 0.2659, -0.6291, 0.6039)^T$$

The first eigenvalue is $e_{y1}=0.0035680$ tells us the about $\frac{e_{y1}}{\sum_{i=1}^5e_{yi}}\approx96%$ of total variance is captured by the associated eigenvector, and the associated eigenvector implies the underlying trend of log-returns of yield. 

```{r, echo=F, message=F, warning=F, include=F}
eigen(cov_ytm)
```

\
The eigenvalues for covariance matrix of log forward rate are $e_{f1}=0.00187295$, $e_{f2}=0.00006053$, $e_{f3}=0.00001179$, and $e_{f4}=0.00000436$, and the eigenvectors are:
$$v_{f1}=(0.4492, 0.5416, 0.5097, 0.4951)^T,v_{f2}=(0.7347, 0.2102, -0.5600, -0.3200)^T$$
$$v_{f3}=(-0.4319, 0.7831, -0.0173, -0.4471)^T, v_{f4}=(-0.2681, 0.2218, -0.6529, 0.6728)^T$$

The first eigenvalue is $e_{f1}=0.00187295$ tells us the about $\frac{e_{f1}}{\sum_{i=1}^4e_{fi}}\approx96%$ of total variance is captured by the associated eigenvector, and the associated eigenvector implies the underlying trend of log-returns of 1-year forward rate.

```{r, echo=F, message=F, warning=F, include=F}
eigen(cov_fwd)
```


## References and GitHub Link to Code

